---
title: "Data analysis for single dose treatment data"
author: "Dina Schuster"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data analysis for single dose treatment data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  <style>
  body {
    text-align: justify}
</style>
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette will give you an overview of how you can analyse single dose treatment LiP-MS or bottom-up proteomics data using **protti**. 
Before analysing your data please make sure that it is of sufficient quality and that you do not have any outliers in your data.
To do this you can take a look at the [quality control vignette](https://jpquast.github.io/protti/articles/quality_control_workflow.html).

## protti

**protti** includes several functions that make it easy for the user to analyse and interpret data from LiP-MS or bottom-up proteomics experiments. The R package includes functions for 

* Quality control (see [quality control vignette](https://jpquast.github.io/protti/articles/quality_control_workflow.html) for more detailed information)
* Data preparation
  + Median normalisation
  + Protein abundance calculation from precursor or peptide intensities
  + Fetching of database information (ChEBI, GO, KEGG, MobiDB, UniProt)
  + Calculation of sequence coverage
* Data analysis
  + Statistical hypothesis tests
  + Fitting of dose-response curves (relevant for experiments with several treatment concentrations)
* Data visualisation 
  + Volcano plots
  + Barcode plots (plots that show protein coverage and significant changes projected onto the protein sequence)
  + Wood's plots
  + Profile plots
* Data interpretation
  + GO-term enrichment
  + Network analysis (based on STRING database information)
  + KEGG pathway enrichment

You can read more about specific functions and how to use them by calling e.g. `?median_normalisation` (for the `median_normalisation()` function). Calling `?` followed by the function name will display the function documentation and give you more detailed information about the function. This can be done for any of the functions included in the package.

This document will give you an overview of data preparation, data analysis, data visualisation and data interpretation functions included in **protti**. It will show you how they can be applied to your data. The examples in this file are run on a [published](https://www.nature.com/articles/s41467-020-18071-x) LiP-MS dataset. In the experiment a lysate was treated with different amounts of the drug rapamycin and it could be shown that the target of rapamycin (FKBP1A) could be successfully identified. For this vignette we are using a filtered version of the original data set which includes only the highest treatment concentration and the control (untreated). For simplicity only 50 proteins are included in the dataset.

The data set is produced from the output of [Spectronaut™](https://biognosys.com/shop/spectronaut). However, if you have any other data such as DDA data that was searched with a different search engine you can still apply **protti**'s functions. Just make sure that your data frame contains [tidy data](https://r4ds.had.co.nz/tidy-data.html). That means data should be contained in a long format (e.g. all sample names in one column) rather than a wide format (e.g. each sample name in its own column). You can easily achieve this by using the `pivot_longer()` function from the `tidyr` package. If you are unsure about how your input data is supposed to look like, please use the `create_synthetic_data()` function and compare this to your data. 

The input data should have a similar structure to this example: 
  
  | Sample  |  Replicate |  Peptide  |  Condition | log2(intensity) |
  |:-------:|:----------:|:---------:|:----------:|:----------------:
  | sample1 |      1     |  PEPTIDER |   treated  |     14          |
  | sample1 |      1     |  PEPTI    |   treated  |     16          |
  | sample1 |      1     |  PEPTIDE  |   treated  |     17          |
  | sample2 |      1     |  PEPTIDER |  untreated |     15          |
  | sample2 |      1     |  PEPTI    |  untreated |     18          |
  | sample2 |      1     |  PEPTIDE  |  untreated |     12          |

# How to use protti to analyse your data
  
## Getting started
  
  Before we can start analysing our data, we need to load the **protti** package. This is done by using the base R function `library()`. In addition, we are also loading the packages `magrittr` and `dplyr`. Both `magrittr` and `dplyr` are part of the [tidyverse](https://www.tidyverse.org), a collection of packages that provide useful functionalities for data processing and visualisation. If you use many tidyverse packages in your workflow you can easily load all at once by calling `library(tidyverse)`.

```{r setup, message = FALSE, warning = FALSE}
library(protti)
library(magrittr)
library(dplyr)
```

After having loaded the required packages we will load our data set into the R environment.
In order to do this for your data set you can use the function `read_protti()` which both reads the data and also cleans up column names ('.' is replaced with "_" and names are converted to lowercase) to make them R-friendly.

For this example we are going to use our test data set included in the R package data folder. To read in the file we are simply going to use the `utils` function `data()`.

```{r load_data}
# To read in your own data you can use read_protti()
# your_data <- read_protti(filename = "mydata/data.csv")

data("rapamycin_10uM")
```

## Data preparation
  
### Log2 transformation, median normalisation and CV filtering
  
After inspecting the data and performing quality control (see [quality control vignette](https://jpquast.github.io/protti/articles/quality_control_workflow.html) for more information) we will now start to prepare the data for the analysis.

Firstly, we remove decoy hits (false hits). Our example data set contains a column called "eg_is_decoy" that consists of logicals indicating whether or not the peptide is a decoy hit. To remove decoys we will use the `dplyr` function `filter()`.

Next, we are log2 transforming our intensities, then we are median normalising the data. To transform the intensities we are using the `dplyr` function `mutate()` which creates a new column while maintaining the original column. 

Note that we are also using the pipe operator `%>%` included in the R package `magrittr`. `%>%` takes the output of the preceding function and supplies it as the first argument of the following function. Using `%>%` makes code easier to read and follow.

For median normalisation we are using the **protti** function `median_normalisation()`. The function normalises intensities for each run to the median of all runs. This is only necessary if your search algorithm does not already median normalise your intensities. For the example data we have disabled median normalisation in Spectronaut, therefore we need to median normalise now. 
The formula for median normalisation is: 

```{block, type='blackbox'}
$$median ~ normalised ~ intensity  = intensity - median ( run ~ intensity ) + median ( global ~ intensity) $$
```

To ensure that our data includes only good peptide measurements we will also filter our data based on coefficients of variation (CV). In order to do this we are using the function `filter_cv()`. For this example we are retaining peptides with a CV < 25 % in at least one of the two conditions.

The CVs are calculated within the function according to the formula: 

```{block, type='blackbox'}
$$CV = \frac{standard ~ deviation}{mean} * 100$$
```
  
```{r median_normalisation_filtering, message = FALSE, warning = FALSE}
data_normalised <- rapamycin_10uM %>%
  filter(eg_is_decoy == FALSE) %>%
  mutate(intensity_log2 = log2(fg_quantity)) %>%
  median_normalisation(sample = r_file_name, 
                       intensity_log2 = intensity_log2)

data_filtered <- data_normalised %>%
  filter_cv(grouping = eg_precursor_id, 
            condition = r_condition, 
            log2_intensity = intensity_log2, 
            cv_limit = 0.25,
            min_conditions = 1)
```

### Remove non-proteotypic peptides

For LiP-MS analysis we commonly remove non-proteotypic peptides (i.e. peptides that could come from more than one protein). If you are using Spectronaut output you will find a column called "pep_is_proteotypic". This column contains logicals indicating whether your peptide is proteotypic or not. This step is optional.

To filter out non proteotypic peptides we are using the `dplyr` function `filter()`.

```{r data_preparation_prot_pep, message = FALSE, warning = FALSE}
data_filtered_proteotypic <- data_filtered %>%
  filter(pep_is_proteotypic == TRUE)
```

### Fetching database information and assigning peptide types

In order to obtain more information about our identified proteins we are going to use the function `fetch_uniprot()` to download some of the available information from UniProt directly. 

`fetch_uniprot()` uses a vector of UniProt IDs as its input. We are producing this vector by using the base R function `unique()` which will extract all unique elements in the selected protein ID column. In this case we want to download the full protein name, gene IDs, GO terms associated with molecular function, StringDB IDs, information on known interactors, location of the active site, location of binding sites, PDB entries, protein length and protein sequence. There are more options for columns to add (for more information on possible columns to add click [here](https://www.uniprot.org/help/uniprotkb_column_names)). 

`fetch_uniprot()` returns a new data frame, in order to be able to merge this with our original data frame we have to rename the ID column to match the name of the protein ID column of our original data frame. To do this we are using `dplyr`'s `rename()` function. 

To merge the two data frames we are using the `dplyr` function `left_join()`. We are matching the two data frames by the column "pg_protein_accessions". By using `left_join()` we are retaining all rows from our original data frame while adding the columns from the data fame generated with `fetch_uniprot()`.

Next, we would like to assign the trypticity of our peptides (i.e. if the peptides are fully-tryptic, semi-tryptic or non-tryptic). In order to do this we will first of all need to define the peptide positions in the protein and find the preceding and following amino acids. To obtain this information we are using `find_peptide()`. The output of the function can then be used in the function `peptide_type()` which will add an additional column with the peptide trypticity information. By using the function `sequence_coverage()` we are adding an additional column to the data frame containing information on how much of the protein sequence we identified in our experiment. 

```{r fetching_database_information, message = FALSE, warning = FALSE}
uniprot_ids <-
  unique(data_filtered_proteotypic$pg_protein_accessions)

uniprot <-
  fetch_uniprot(
    uniprot_ids = uniprot_ids,
    columns = c(
      "protein names",
      "genes",
      "go(molecular function)",
      "database(String)",
      "interactor",
      "feature(ACTIVE SITE)",
      "feature(BINDING SITE)",
      "database(PDB)",
      "length",
      "sequence"
    )
  ) %>%
  rename(pg_protein_accessions = id)

data_filtered_uniprot <- data_filtered_proteotypic %>%
  left_join(y = uniprot,
            by = "pg_protein_accessions") %>%
  find_peptide(protein_sequence = sequence,
               peptide_sequence = pep_stripped_sequence) %>%
  peptide_type(aa_before = aa_before,
               last_aa = last_aa) %>%
  sequence_coverage(protein_sequence = sequence,
                    peptides = pep_stripped_sequence)
```

## Data analysis

### Statistical hypothesis test

To test if there is a difference between the peptide abundances in our two conditions (i.e. rapamycin treated and untreated) we are going to use Welch's t-test (unpaired). Welch's t-test is a method for comparing means of two independent sample groups and it does not assume equal variances between groups (as Student's t-test would).

Before the statistical hypothesis thest we have to define the types of missing values present in our data set. We are going to use the function `assign_missingness()` which will return a column with information on the types of missingness we have in our data (i.e. complete, missing at random (MAR) or missing not at random (MNAR)). We are using the default parameters of this function which assumes that missing values are MAR when the conditions are at least 70 % complete. Missing values are assumed to be MNAR when less than 25 % of values are present in one condition if the other condition is complete. 

After assigning the types of data missingness we are using the function `diff_abundance()`. In this case by selecting `method = t-test` the function will perform a Welch's t-test. There are also options included to perform a moderated t-test based on the R package `limma` or to detect differential abundances based on the algorithm implemented in the R package `proDA`. The algorithm used for `proDA` is based on a probabilistic dropout model which facilitates hypothesis testing without imputation. 

Please note that in this case we are not imputing missing values. You can, however, do so by using the function `impute()`. This function uses the output of `assign_missingness()` as its input. You can use two different imputation methods: 

* `method = ludovic` will sample values that are MNAR from a normal distribution around a value that is 3 (log2) lower than the lowest intensity in your data set. The method is named after and created by our very helpful and appreciated colleague [Ludovic Gillet](https://imsb.ethz.ch/research/picotti/PeoplePicotti/ludovicgillet.html).
* `method = noise` will sample MNAR values from a normal distribution around the mean noise. This requires you to have an additional column with information on the noise.

As we are dealing with a LiP-MS data set we are performing the statistical analysis on precursor level. For protein abundance data you can simply use protein abundances as your intensities and select your protein groups column for the grouping argument. 
After performing the t-test we are rejoining the output with our previous data frame to ensure that we have all the information in one data frame. 

```{r t_test, message = FALSE, warning = FALSE}
diff_abundance <- data_filtered_uniprot %>%
  assign_missingness(
    sample = r_file_name,
    condition = r_condition,
    grouping = eg_precursor_id,
    intensity = normalised_intensity_log2,
    ref_condition = "control",
    completeness_MAR = 0.7,
    completeness_MNAR = 0.25
  ) %>%
  diff_abundance(
    sample = r_file_name,
    condition = r_condition,
    grouping = eg_precursor_id,
    intensity = normalised_intensity_log2,
    missingness = missingness,
    comparison = comparison,
    ref_condition = "control",
    method = "t-test"
  ) %>%
  left_join(y = data_filtered_uniprot, by = "eg_precursor_id")
```

### Volcano plot

Next we are going to visualise the output of the previously performed hypothesis test to assess the resuls of our experiment. For this we are going to plot a volcano plot with fold-changes on the x-axis and the adjusted p-value (q-value) on the y-axis. 
By adjusting our p-values for multiple testing we are ensuring that we keep the false discovery rate low. 
The output of the previously used `diff_abundance()` function is ideal to use for the `volcano_protti()` function as it contains all the information we need: precursor IDs, protein IDs, fold changes (`diff`) and adjusted p-values (`adj_pval`).
We are going to highlight the peptides of the known target of rapamycin FKBP1A (UniProt ID = P62942) in blue to quickly find the peptides in the plot. You can also make the plot interactive by setting `interactive = TRUE`. This will help you quickly obtain more information on each point in the plot.
We are going to plot the volcano plot with and without adjusted p-values. 

Note: If very little observations are changing significantly or if the p-values are not very low to begin with it can happen that only a few significant hits remain after adjusting the p-value.

```{r volcano_plot, fig.align= "center", fig.width = 6, fig.height = 5, message = FALSE, warning = FALSE}
volcano_protti(
  data = diff_abundance,
  grouping = eg_precursor_id,
  log2FC = diff,
  significance = pval,
  method = "target",
  target_column = pg_protein_accessions,
  target = "P62942",
  x_axis_label = "log2(fold change) rapamycin treated vs. untreated",
  y_axis_label = "-log10(p-value)",
  title = "Volcano plot (p-value not adjusted)"
)

volcano_protti(
  data = diff_abundance,
  grouping = eg_precursor_id,
  log2FC = diff,
  significance = adj_pval,
  method = "target",
  target_column = pg_protein_accessions,
  target = "P62942",
  x_axis_label = "log2(fold change) rapamycin treated vs. untreated",
  y_axis_label = "-log10(q-value)",
  title = "Volcano plot"
)
```

### Barcode plot

For LiP-MS experiments a good way to see where on the protein the changes due to binding or conformational changes occur is to plot a barcode plot. In a barcode plot plotted with `barcode_plot()` the detected peptides are coloured in grey and the changing peptides are highlighted in blue.

In order to produce a barcode plot only for our target FKBP1A we are creating a dataframe that contains only information for our target protein. We are doing this by using `dplyr`'s `filter()` function. The filtered data frame is then used as the input for the plot.

```{r barcode_plot, fig.align = "center", fig.width = 6, message = FALSE, warning = FALSE}
FKBP12 <- diff_abundance %>%
  filter(pg_protein_accessions == "P62942")

barcode_plot(
  data = FKBP12,
  start_position = start,
  end_position = end,
  protein_length = length,
  coverage = coverage,
  colouring = diff,
  cutoffs = c(diff = 1, adj_pval = 0.01),
  protein_id = pg_protein_accessions
  )
```

### Wood's plot

An additional way to plot LiP-MS changes are Wood's plots. These plots will show the extent of the change along the protein sequence.
To produce a Wood's plot we are using the function `woods_plot()` and colouring the peptides according to their adjusted p-values. 

```{r woods_plot, fig.align = "center", fig.width = 6, message = FALSE, warning = FALSE}
woods_plot(
  data = FKBP12,
  fold_change = diff,
  start_position = start,
  end_position = end,
  protein_length = length,
  coverage = coverage,
  colouring = adj_pval,
  protein_id = pg_protein_accessions
  )
```

### Peptide profile plots

To see how the individual precursors in our target protein are changing with the treatment we are plotting profile plots. We are using the function `plot_peptide_profiles()` to generate the following plot.

```{r protile_plot, fig.align = "center", fig.width = 10, fig.height = 6, message = FALSE, warning = FALSE}
plot_peptide_profiles(
  data = FKBP12,
  sample = r_file_name,
  peptide = eg_precursor_id,
  intensity = normalised_intensity_log2,
  grouping = pg_protein_accessions,
  targets = "P62942",
  split_all = FALSE,
  protein_abundance_plot = FALSE
)
```

## Additional helpful functions

**protti** includes additional helpful functions that do not make sense to use for this data set but are great to use for data sets with systematic changes. These functions include the `go_enrichment()`function that helps you check if your hits are enriched for specific gene ontology (GO) terms, the `network_analysis()` function that plots a String network based on information from StringDB for your hits and the `kegg_enrichment()` function which checks for enriched pathways in your hits.

Here we will quickly show you how you could use these functions in a data set with more global changes:

For GO enrichment you would add an additional column to your data frame containing information on whether your hit is significant or not. You can do this by using the `dplyr` function `mutate()` which will add an additional column to your data frame. Here we want the column to contain logicals that are either TRUE when the adjusted p-value is below 0.01 and the log2(fold change) is below -1 or above 1 or to be FALSE if this is not the case. We are using the `data.table` function `ifelse()` to produce the logicals.

For the network analyis we are filtering the previously produced data frame containing the `is_significant` column for significant hits. This data frame can then be used as an input for `network_analysis()` to check if the proteins can be found in an interaction network based on StringDB information. 

For `kegg_enrichment()` you need to first use the function `fetch_kegg()` to obtain the KEGG pathway identifiers for your data set. You can then use `dplyr`'s `right_join()`to join the output with the previously produced data frame containing a column indicating whether your hits are significant or not. When joining the data frames make sure that the protein ID columns by which you are joining your data frames have the same name. Now you are ready to use `kegg_enrichment()`. 

```{r additional_functions}
# go_df <- diff_abundance %>%
#   mutate(is_significant = ifelse(
#     test = (adj_pval < 0.01 & abs(diff) > 1),
#     yes = TRUE,
#     no = FALSE
#   ))
# 
# go_enrichment(
#   data = go_df,
#   protein_id = pg_protein_accessions,
#   is_significant = is_significant,
#   go_annotations_uniprot = go_molecular_function
# )
# 
# network_df <- go_df %>%
#   filter(is_significant == TRUE)
# 
# network_analysis(data = network_df, 
#                  protein_id = pg_protein_accessions, 
#                  string_id = database_string, 
#                  organism_id = 9606)
# 
# kegg_df <- fetch_kegg(species = "hsa") %>%
#   rename(pg_protein_accessions = uniprot_id) %>%
#   right_join(y = go_df, by = "pg_protein_accessions")
# 
# kegg_enrichment(data = kegg_df,
#                 protein_id = pg_protein_accessions, 
#                 is_significant = is_significant,
#                 pathway_id = pathway_id, 
#                 pathway_name = pathway_name)
```

