---
title: "Qualtiy control (QC) workflow"
author: "Dina Schuster"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Qualtiy control (QC) workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style>
body {
text-align: justify}
</style>

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

When working with proteomics and limited proteolysis-coupled mass spectrometry ([LiP-MS](https://doi.org/10.1038/nprot.2017.100)) data, it is important to ensure that the data is of sufficient quality before inferring biological relevance and meaning from it. In general, there are several ways in which this can be done: 

  * Inspection of .raw files acquired directly on the mass spectrometer (e.g. using vendor specific applications to visualize mass spectrometry data, such as the Thermo Scientific applications XCalibur™ and FreeStyle™)
  * Different search engines offer quality control functionalities (e.g. [Spectronaut™](https://biognosys.com/shop/spectronaut) offers a lot of QC options)
  * R packages (e.g. [PTXQC](https://doi.org/10.1021/acs.jproteome.5b00780) which runs on [MaxQuant](https://www.maxquant.org/) output or [rawDiag](https://doi.org/10.1021/acs.jproteome.8b00173) which can be used directly on raw files) 

_Please note that this is not a complete list, it is just intended to give you an idea of the different possibilities. There are many more software options and R packages published to date._

The idea behind all of the options listed above is to assess if the data quality is homogenous among the different measurements ( = samples) of an experiment and if there are any outliers that should be excluded. 

# protti

**protti** includes several functions that make it easy for the user to compare the different samples in one experiment. Some functions in this R package are specific for LiP-MS data analyses, whereas others can also be used for general proteomics analyses. 

You can read more about specific functions and how to use them by calling e.g. `?qc_cvs` (for the `qc_cvs()` function). Calling `?` followed by the function name will display the function documentation and give you more detailed information about the function. This can be done for any of the functions included in the package.

This document will give you an overview of QC functions included in **protti** and will show you how they can be applied to your data. The examples in this file are run on synthetic data, which is created with the function `create_synthetic_data()`.

The functions in **protti** are tailored towards the analysis of DIA data, based on the output of the search engine [Spectronaut™](https://biognosys.com/shop/spectronaut). However, if you have any other data, such as DDA data that was searched with a different search engine, you can still apply our functions. Just make sure that your data frame contains [tidy data](https://r4ds.had.co.nz/tidy-data.html). That means data should be contained in a long format (e.g. all sample names in one column) rather than a wide format (e.g. each sample name in its own column). You can easily achieve this by using the `pivot_longer()` function from the `tidyr` package. If you are unsure about how your input data is supposed to look like, please use the `create_synthetic_data()` function and compare this to your data. 

The input data should have a similar structure to this example: 

| Sample  |  Replicate |  Peptide  |  Condition | log2(intensity) |
|:-------:|:----------:|:---------:|:----------:|:----------------:
| sample1 |      1     |  PEPTIDER |   treated  |     14          |
| sample1 |      1     |  PEPTI    |   treated  |     16          |
| sample1 |      1     |  PEPTIDE  |   treated  |     17          |
| sample2 |      1     |  PEPTIDER |  untreated |     15          |
| sample2 |      1     |  PEPTI    |  untreated |     18          |
| sample2 |      1     |  PEPTIDE  |  untreated |     12          |

## Getting started

Before we can start analysing our data, we need to load the **protti** package. This is done by using the base R function `library()`. In addition, we are also loading the packages `magrittr` and `dplyr`. Both `magrittr` and `dplyr` are part of the [tidyverse](https://www.tidyverse.org), a collection of packages that provide useful functionalities for data processing and visualisation. If you use many tidyverse packages in your workflow you can easily load all at once by calling `library(tidyverse)`.

```{r setup, message = FALSE, warning = FALSE}
library(protti)
library(magrittr)
library(dplyr)

```

After having loaded the required packages we can create a synthetic dataset, that contains data similar to data obtained from a treatment experiment with e.g. a protein, metabolite or small-molecule.
You can skip this step if you have your own data ready.

We are creating a random dataset with 100 different proteins, out of which 5 % are significantly changing upon treatment. The dataset includes 3 replicates for 2 different conditions (treated and untreated).

The output of the `create_synthetic_data()` function is modeled after real LiP-MS data and its commonly observed data distributions.

Please note that generally, quality control should be conducted on raw unfiltered data (the direct output of your search engine of choice).

```{r create_synthetic_data}
set.seed(123) # by setting the seed we are making sure that the random object generation can be reproduced

data <- create_synthetic_data(n_proteins = 100,
                              frac_change = 0.05,
                              n_replicates = 3,
                              n_conditions = 2,
                              method = "random_effect",
                              additional_metadata = TRUE)

```

## Quality control

### Coefficients of variation

A good first step of quality control is to check if coefficients of variation between replicates are in a reasonable range. Ideally, these should be below 15-20 %. If you see groups with higher CVs, there might have been a sample preparation error or the instrument performance was not ideal at the time of measurement. You can check if after `median_normalisation()` (depending on whether your search engine has performed median normalisation already) your CVs improve.

To conduct the CV based quality control, we are making use of the function `qc_cvs()`. This function calculates the coefficients of variation. The output can either be a table or a plot of coefficients of variations for each condition. Below we are returning a CV table and a violin plot, showing the CVs for different conditions.

The CVs (in percent) are calculated with the following formula: 

```{block, type='blackbox'}

$$CV = \frac{standard deviation}{mean} * 100$$
```


Since the CV function works only with raw values, we will backtransform our log2 transformed data and create a new column called `raw_intensity` from the `peptide_intensity_missing` column which contains peptide intensities. To create the new column, we are using the function `mutate()` from the R package `dplyr`.
We also make use of the pipe operator `%>%` included in the R package `magrittr` which takes the output of the preceding function and supplies it as the first argument of the following function. Using `%>%` makes code easier to read and follow. 

The "combined" group of CVs contains CVs across all samples and not only across the replicates of a certain condition. Ideally the combined CVs should always be the group with the highest CV.  

```{r qc_cvs, fig.width = 6, fig.height = 4, fig.align = "center"}
input <- data %>%
  mutate(raw_intensity = 2^peptide_intensity_missing) # as the data is log2 transformed, we need to backtransform it before calculating the CVs

qc_cvs(data = input,
       grouping = peptide,
       condition = condition,
       intensity = raw_intensity, 
       plot = FALSE)

qc_cvs(data = input,
       grouping = peptide,
       condition = condition,
       intensity = raw_intensity, 
       plot = TRUE, 
       plot_style = "violin")
```

### Number of identifications (IDs)

The number of protein or peptide identifications should be similar for different samples. If there are significanlty less observations in one sample, this might indicate a sample preparation or measurement error.

For the analysis of the number of identifications of precursors, peptides or proteins we are using the function `qc_ids()`. This function can return either a table or a plot. The output of this function - and also of a lot of other **protti** functions - can be plotted in an interactive plot that makes use of the R package `plotly`. You can plot an interactive version of the plot by setting `interactive = TRUE` within the function call. 

```{r qc_ids, fig.width = 6, fig.height = 4, fig.align = "center"}
qc_ids(data = input,
       sample = sample,
       grouping = protein,
       intensity = peptide_intensity_missing,
       condition = condition, 
       plot = FALSE)

qc_ids(data = input,
       sample = sample,
       grouping = protein,
       intensity = peptide_intensity_missing,
       condition = condition, 
       title = "Protein identifications per sample",
       plot = TRUE)
```

### Peptide types

Now that we had a look at protein IDs, we should also check peptide identifications for homogeneity across samples. An important factor to take into account when analysing LiP-MS data are peptide types (tryptic, semi-tryptic, non-tryptic). They can give us an idea about the reproducibility of cleavage events in the experiment and if our treatment influenced protease activity or the digest in general. 

In a standard LiP-MS experiment we use 3 different proteases: 

  * Proteinase K (unspecific, cleaves accessible protein regions)
  * Trypsin (specific, cleaves after lysines or arginines)
  * Lys-C (specific, cleaves after lysines)

Peptide types can be assessed in Spectronaut directly or by using the **protti** function `peptide_type()`. This function takes amino acids before and after the first and last amino acid in the peptide into account. A peptide that starts after a lysine or arginine and ends with a lysine or arginine is a _tryptic_ peptide. A peptide that has one proteinase K cleavage site (not lysine or arginine) and one trypsin cleavage site either at the beginning or at the end of the peptide is a _semi-tryptic_ peptide. A peptide that does not have any trypsin cleavage sites is a _non-tryptic_ peptide. If information about the amino acid before and the last amino acid of a peptide are not available in your data you can use the `find_peptide` function from the **protti** package to create it.

We are going to use the function `qc_peptide_type()` to evaluate the distribution of peptide types between our samples. The function can return a plot or a table, based on either peptide type counts or intensities. Both are equally important when assessing how homogenous the samples are. For `method = intensity` the function uses raw intensity values, so we are going to use the column `raw_intensity` which we added previously. 

In a typical LiP-MS experiment, there should be little non-tryptic peptides. The peptide type distribution depends on your digestion conditions (i.e. temperature, duration, concentration).

```{r qc_peptide_type, fig.width = 6, fig.height = 4, fig.align = "center"}
qc_peptide_type(data = input,
                sample = sample, 
                peptide = peptide, 
                pep_type = pep_type, 
                method = "intensity", 
                intensity = raw_intensity, 
                plot = TRUE, 
                interactive = FALSE)

qc_peptide_type(data = input,
                sample = sample, 
                peptide = peptide, 
                pep_type = pep_type, 
                method = "count", 
                plot = TRUE, 
                interactive = FALSE)
```

### Run intensities

The function `qc_intensity_distribution()` plots all precursor, peptide or protein intensities for each sample as a boxplot if method "boxplot" is selected. This is helpful for quick assessment of any major sample losses or measurement issues. 

A similar way in which the run intensities can be assessed is to plot the median run intensities as a line plot. This helps you quickly assess if there are any trends in your data. The function to use for this analysis is `qc_median_intensities()`. 

```{r qc_intensity_distribution_boxplot, fig.width = 6, fig.height = 4, fig.align = "center"}

qc_intensity_distribution(data = input,
                          sample = sample,
                          grouping = peptide,
                          intensity = peptide_intensity_missing,
                          plot_style = "boxplot")

qc_median_intensities(data = input,
                      sample = sample,
                      grouping = peptide,
                      intensity = peptide_intensity_missing)
```

### Charge states

The charge state distibution of the detected peptides can be assessed with `qc_charge_states()`. Similar to the peptide types, this should of course also be homogenous for the different samples. The function can return either a plot or a table based on counts of peptides with a specific charge state or the intensities of these peptides. 
For `method = "intensity"` the function requires the raw intensity values created previously as its input.

```{r qc_charge_states, fig.width = 6, fig.height = 4, fig.align = "center"}
qc_charge_states(data = input, 
                 sample = sample, 
                 grouping = peptide, 
                 charge_states = charge, 
                 method = "intensity",
                 intensity = raw_intensity, 
                 plot = TRUE)
```

### Missed cleavages

An important measure of efficient digests with e.g. trypsin are the number of missed cleavages. The proteases trypsin and Lys-C both cleave after positively charged amino acids lysine (K) and arginine (R). If a measured peptide includes either K or R in its sequence, not at the last position, it contains a missed cleavage site. 
These peptides are important to take into account because they can tell you if your tryptic digest was inefficient. 
The number of missed cleavages should generally be low in a proteomics or LiP-MS dataset. 

We are going to check the numbers of missed cleavages in our dataset by using the function `qc_missed_cleavages()`. It can assess missed cleavages based on the count of peptides with missed cleavages or the intensities of the corresponding peptides. For `method = "intensity"` the function uses the raw (not log2 transformed) intensity values. You can have the function either return a plot or a table.  

```{r qc_missed_cleavages, fig.width = 6, fig.height = 4, fig.align = "center"}
qc_missed_cleavages(data = input, 
                    sample = sample, 
                    grouping = peptide, 
                    missed_cleavages = n_missed_cleavage, 
                    method = "intensity",
                    intensity = raw_intensity, 
                    plot = TRUE)
```

### Sequence coverage

The following function gives you insight into the protein coverage percentages in the dataset. For a LiP-MS experiment high sequence coverages are desirable. 

To assess the protein coverage distribution, we are going to use the function `qc_protein_coverage()`. If you do not have a column containing the protein coverages in your data, you can use the function `sequence_coverage()` to obtain this information. 

```{r qc_sequence_coverage, fig.width = 6, fig.height = 4, fig.align = "center"}
qc_sequence_coverage(data = input, 
                     protein_identifier = protein, 
                     coverage = coverage)
``` 

### Peak width

In order to identify potential chromatographic issues that might have occurred during the measurement we can have a look at the chromatographic median peak widths over the complete chromatogram. To do this we are going to use the function `qc_peak_width()`. This function requires either the peak start and end times or the retention time and peak width. This information can be obtained from Spectronaut (or any other search engine of your choice). 
The peak widths should be similar for all the measured samples of the experiment. 

```{r qc_peak_width, fig.width = 6, fig.height = 4, fig.align = "center"}
qc_peak_width(data = input,
              sample = sample, 
              intensity = peptide_intensity_missing,
              retention_time = retention_time, 
              peak_width = peak_width)
``` 

### Data completeness

he following function `qc_data_completeness()` checks how many of all detected precursors, peptides or proteins were identified in each sample. The function can return either a plot or a table. 

```{r qc_data_completeness, fig.width = 6, fig.height = 4, fig.align = "center"}
qc_data_completeness(data = input, 
                     sample = sample, 
                     grouping = peptide, 
                     intensity = peptide_intensity_missing, 
                     plot = TRUE)
``` 

### Log2 Intensity distribution

For different kinds of analyses (e.g. t-tests) it is important that your data intensity follows a normal distribution. To ensure that this is the case, we are going to use the function `qc_intensity_distriubution()`. The function returns a histogram plot when `plot_style = "histogram"` showing how the intensities are distributed.

```{r qc_intensity_distribution_histogram, fig.width = 6, fig.height = 4, fig.align = "center"}
qc_intensity_distribution(data = input,
                          grouping = peptide,
                          intensity = peptide_intensity_missing,
                          plot_style = "histogram")
``` 

### Sample correlation

Another way of quality control is to check the correlation of your samples. Ideally, replicates should cluster together and different treatment conditions should be separated. We are now going to check if this is the case for our data by using the function `qc_sample_correlation()`. The function will return a correlation heatmap with a comparison of all samples.

```{r qc_sample_correlation, fig.width = 6, fig.height = 4, fig.align = "center"}
 qc_sample_correlation(data = input,
                       sample = sample, 
                       grouping = peptide, 
                       intensity = peptide_intensity_missing, 
                       condition = condition, 
                       interactive = FALSE)
``` 

### Principal component analysis (PCA)

Another popular quality control method (which could also be considered part of data analysis) is the principal component analysis (PCA). Generally, PCA is a method that reduces dimensionality of large datasets. In our case this helps us to quickly assess how similar or different our replicates and conditions are. We are going to use `qc_pca()` to compute and plot a PCA for our data.

```{r qc_pca, fig.width = 6, fig.height = 4, fig.align = "center"}
 qc_pca(data = data,
        sample = sample, 
        grouping = peptide, 
        intensity = peptide_intensity_missing, 
        condition = condition)
``` 

### Additional QC functions

There are two additional QC functions that are not included in this vignette: 

* `qc_contaminants()` is a function that can be used if proteins are assigned as contaminants. This information can e.g. be exported from MaxQuant. The function calculates the percentage of proteins annotated to be contaminants and returns either a table or a plot.
* `qc_proteome_coverage()` calculates the proteome coverage for each sample and the whole experiment and returns either a plot or a table.
